
# Pictures of Stress {#picsstress}

Even for $n$ as small as four and $p$ as small as two the dimension of
the space of centered configurations is six, and there is no natural way
to visualize a function of six variables. What we can do is plot stress
on two-dimensional subspaces, either as a contour plot or as a perspective plot. Our two-dimensional subspaces are of the form $\alpha X+\beta Y$, where $X$ and $Y$ are fixed configurations. Much of
this chapter is a modified, and in some places expanded, version of @deleeuw_E_16l.

```{r example, echo = FALSE}
delta <- as.matrix (dist (diag (4)))
delta <- delta * sqrt (2 / sum (delta ^ 2))
w <- 1 - diag(4)
x <- matrix (c(0,0,1,0,0,1,1,1), 4, 2, byrow = TRUE)
y <- matrix (c(0,0,1,0,.5,.5*sqrt(3)), 3, 2, byrow = TRUE)
y <- rbind(y, colSums(y) / 3)
x <- apply (x, 2, function (z) z - mean (z))
y <- apply (y, 2, function (z) z - mean (z))
d1 <- as.matrix (dist (x))
d2 <- as.matrix (dist (y))
lbd1 <- sum (delta * d1) / sum (d1 ^ 2)
x <- x * lbd1
d1 <- d1 * lbd1
lbd2 <- sum (delta * d2) / sum (d2 ^ 2)
y <- y * lbd2
d2 <- d2 * lbd2
sx<-sum((delta-d1)^2)/2
sy<-sum((delta-d2)^2)/2
```

Throughout we use a small example of order $n=4$ which all
dissimilarities equal. The same example has been analyzed by
@deleeuw_A_88b, @deleeuw_R_93c, @trosset_mathar_97, and
@zilinskas_podlipskyte_03. For this example a global minimum in two
dimensions has its four points in the corners of a square. That is our
$X$, which has stress `r sx`. Our $Y$ is another stationary point, which
has three points in the corners of an equilateral triangle and the
fourth point in the center of the triangle. Its stress is `r sy`. We
column-center the configurations and scale them so that they are
actually stationary points, i.e. so that $\eta^2(X)=\rho(X)$ and
$\eta^2(Y)=\rho(Y)$. The example is chosen in such a way that there are non-zero $\alpha$ and $\beta$ such that $d_{12}(\alpha X+\beta Y)=0$. In fact $d_{12}$ is the only distance that can be made zero by a non-trivial linear combination.

Another way of looking at the two configurations is that $X$ are four
points equally spaced on a circle, and $Y$ are three points equally
spaced on a circle with the fourth point in the center of the circle.
@deleeuw_A_88b erroneously claims that $Y$ is a non-isolated local
minimum of stress, but @trosset_mathar_97 have shown there exists a
descent direction at $Y$, and thus $Y$ is actually a saddle point. Of
course the stationary points defined by $X$ and $Y$ are far from unique,
because we can permute the four points over the corners of the square
and the triangle in many ways.

## Coefficient Space

Configurations as a linear combination of a number of given
configurations have already been discussed in general in chapter
\@ref(propchapter), section \@ref(propspaces) as the transformation from
configuration space to coefficient space. Since we are dealing here with
the special case of linear combinations of only two configurations we
specialize some of these general results.

We start with $d_{ij}^(\theta)=\theta'T_{ij}\theta$, where $\theta$ has
elements $\alpha$ and $\beta$, and where $T$ is the $2\times 2$ matrix
with elements \begin{equation}
t_{ij}:=\begin{bmatrix}
\text{tr}\ X'A_{ij}X&\text{tr}\ X'A_{ij}Y\\
\text{tr}\ Y'A_{ij}X&\text{tr}\ Y'A_{ij}Y
\end{bmatrix}
(\#eq:pictbform)
\end{equation}

Then

\begin{equation}
\tilde\sigma(\theta):=1-2\ \theta'C(\theta)\theta+\theta'U\theta,
(\#eq:picstress2)
\end{equation}

where, using $Z(\theta)=\alpha X+\beta Y$,

\begin{equation}
C(\theta):=
\begin{bmatrix}
\text{tr}\ X'B(Z(\theta))X&\text{tr}\ X'B(Z(\theta))Y\\
\text{tr}\ Y'B(Z(\theta))X&\text{tr}\ Y'B(Z(\theta))Y
\end{bmatrix},
(\#eq:pictbformc)
\end{equation}

and

\begin{equation}
U:=\begin{bmatrix}
\text{tr}\ X'VX&\text{tr}\ X'VY\\
\text{tr}\ Y'VX&\text{tr}\ Y'VY
\end{bmatrix}.
(\#eq:pictbformu)
\end{equation}

We have used $\tilde\sigma$ in equation \@ref(eq:picstress2) to
distinguish stress on the two-dimensional space of coefficients from
stress on the eight-dimensional space of $4\times 2$ configurations.
Thus $\tilde\sigma(\alpha,\beta)=\sigma(\alpha X + \beta Z)$.

The gradient at $\theta$ is

\begin{equation}
\nabla\tilde\sigma(\theta)=U\theta-C(\theta)\theta,
(\#eq:picgrad2)
\end{equation}

and the Hessian is

\begin{equation}
\nabla^2\tilde\sigma(\theta)=U-\mathop{\sum\sum}_{1\leq i<j\leq n}
w_{ij}\frac{\delta_{ij}}{d_{ij}(\theta)}\left\{T_{ij}-\frac{T_{ij}\theta\theta'T_{ij}}{\theta'T_{ij}\theta}\right\}.
(\#eq:pichess2)
\end{equation}

------------------------------------------------------------------------

```{theorem, label = "pictpreserve"}
If $B(X)X=VX$ and $\theta=\begin{bmatrix}1&0\end{bmatrix}$ then  $C(\theta)\theta=U\theta$.
```

------------------------------------------------------------------------

```{proof}
If $B(X)X=VX$ and $\theta=\begin{bmatrix}1&0\end{bmatrix}$ then, by equations \@ref(eq:pictbformc) and \@ref(eq:pictbformu),

\begin{equation}
U-C(\theta)=\begin{bmatrix}0&0\\0&\text{tr}\ Y'(V-B(X))Y\end{bmatrix}.
(\#eq:picuminc)
\end{equation}

Thus $(U-C(\theta))\theta=0$.
```

------------------------------------------------------------------------

Thus each stationary point of $\sigma$ gives a stationary point of
$\tilde\sigma$. The other way around, however, we are not so lucky.

------------------------------------------------------------------------

```{theorem, label = "pictaround"}
If $C(\theta)\theta=U\theta$ and if the $n\times 2p$ matrix
$\begin{bmatrix}X&Y\end{bmatrix}$ has rank $n-1$ then  $B(Z)Z=VZ$.
```

------------------------------------------------------------------------

```{proof}
If  $C(\theta)\theta=U\theta$ then both $\text{tr}\ X'(V-B(Z)Z)=0$
and $\text{tr}\ Y'(V-B(Z))Z=0$. If the $n\times 2p$ matrix
$\begin{bmatrix}X&Y\end{bmatrix}$ has rank $n-1$ then this implies
$(V-B(Z))Z=0$.
```

------------------------------------------------------------------------

In our example the singular values of $\begin{bmatrix}X&Y\end{bmatrix}$
are `r svd(cbind(x,y))$d` and thus there is a one-one correspondence
between stationary points of $\sigma$ and $\tilde\sigma$.

------------------------------------------------------------------------

```{theorem, label = "picsecder"}
If $B(X)X=VX$ and $\theta=\begin{bmatrix}1&0\end{bmatrix}$ then 

1. If $\text{tr}\ Y'(V-B(X))Y > 0$ then $\sigma$ has a local minimum at theta.
2. If $\sigma$ has a saddle point at $\theta$ then $\text{tr}\ Y'(V-B(X))Y < 0$.

```

------------------------------------------------------------------------

```{proof}
Suppose $B(X)X=VX$ and $\theta=\begin{bmatrix}1&0\end{bmatrix}$. Then,
from \@ref(eq:pichess2)  and \@ref(eq:picuminc),
$$
\nabla^2\sigma(\theta)=\begin{bmatrix}0&0\\0&\text{tr}\ Y'(V-B(X))Y\end{bmatrix}+\mathop{\sum\sum}_{1\leq i<j\leq n}
w_{ij}\frac{\delta_{ij}}{d_{ij}(\theta)}\frac{T_{ij}\theta\theta'T_{ij}}{\theta'T_{ij}\theta}.
$$
```

------------------------------------------------------------------------

In our example $\text{tr}\ X'(V-B(Y))X$ is
`r 4 * sum (x ^ 2) - sum(x * (smacofBmatR(d2, w, delta) %*% x))` and
$\text{tr}\ Y'(V-B(X))Y$ is
`r 4 * sum (y ^ 2) - sum(y * (smacofBmatR(d1, w, delta) %*% y))`.

## Global Perspective

```{r pictmores, echo = FALSE}
by <- - delta/(d2+diag(4))
diag(by)<--rowSums(by)
z <- matrix (0, 2, 2)
z[, 1] <- x[1, ] - x[2, ]
z[, 2] <- y[1, ] - y[2, ]
gm0<-eigen(crossprod(z))$vectors[,2]
xint<-gm0[1]*x+gm0[2]*y
asum <- 2 * 4 * matrix (c (sum(x ^ 2), sum (x * y), sum (x * y), sum (y ^ 2)), 2, 2)
bsum <- chol (asum)
csum <- solve (bsum)
th0 <- bsum %*% gm0
sth1 <- smacof2 (1.0406404, 0.8849253, x, y, delta, eps=1e-15,verbose=FALSE)
sth2 <- smacof2 (1, 0, x, y, delta, eps=1e-15,verbose=FALSE)
sth3 <- smacof2 (0, 1, x, y, delta, eps=1e-15,verbose=FALSE)
th1 <- csum %*% sth1$theta
xso1 <- th1[1] * x + th1[2] * y
th2 <- csum %*% sth2$theta
xso2 <- th2[1] * x + th2[2] * y
th3 <- csum %*% sth3$theta
xso3 <- th3[1] * x + th3[2] * y
sth4 <- newton2(0, 1, x, y, delta, eps=1e-15,verbose = FALSE)
th4 <- csum %*% sth4$theta
xso4 <- th4[1] * x + th4[2] * y
sth5<-newton2(1.12383710,0.77620456,x,y,delta,eps=1e-15,verbose=FALSE)
th5 <- csum %*% sth5$theta
xso5 <- th5[1] * x + th5[2] * y
```

We first make a global perspective plot, over the range $(-2.5,+2.5)$.

```{r first_array, echo = FALSE, cache = TRUE}
aa <- bb <- seq (-2.5, 2.5, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

```{r globalperspective, fig.align= "center", fig.cap = "Global Perspective", fig.width = 10, fig.height = 10, echo = FALSE, cache = TRUE}
par(pty="s")
persp(aa, bb, z, col = "RED", xlab = "theta_1", ylab = "theta_2", zlab = "stress")
```

We see the symmetry, following from the fact that stress is even. We
also see the local maximum at the origin, where stress is not
differentiable. Also note the ridge, where $d_{12}(\theta)=0$ and where
stress is not differentiable either. The ridge shows nicely that on rays
emanating from the origin stress is a convex quadratic. Also, far away
from the origin, stress globally behaves very much like a convex
quadratic (except for the ridge). Clearly local minima must be found in
the valleys surrounding the small mountain at the origin, all within the
sphere with radius $\sqrt{2}$.

## Global Contour

Figure \@ref(fig:globalperspective) is a contour plot of stress over
$(-2,+2)\otimes(-2,+2)$. The red line is
$\{\theta\mid d_{12}(\theta) = 0\}$. The blue line has the minimum of
the convex quadratic on each of the rays through the origin. Thus all
local minima, and in fact all stationary points, are on the blue line.
Note that the plot uses $\theta$ to define the coordinate axes, not
$\gamma=(\alpha,\beta)$. Thus there are no stationary points at $(0,1)$
and $(1,0)$, but at the corresponding points (`r bsum[,1]`) and
(`r bsum[,2]`) in the $\theta$ coordinates (and, of course, at their
mirror images).

Besides the single local maximum at the origin, it turns out that in
this example there are five pairs of stationary points. Or, more
precisely, I have not been able to find more than five. Each stationary
point $\theta$ has a mirror image $-\theta$. Three of the five are local
minima, two are saddle points. Local minima are plotted as blue points,
saddle points as red points.

```{r globalcontour, fig.align= "center", fig.cap = "Global Contour", fig.width = 10, fig.height = 10, echo = FALSE, cache = TRUE}
contour(aa, bb, z, levels = seq(0,.30,length=50))
aa <- seq (-2 * pi,  2 * pi, length = 100)
bb <- sin (aa)
cc <- cos (aa)
xxx <- matrix (0, 100, 2)
for (k in 1:100) {
  z <- c (bb[k], cc[k])
  dd <- matrix (0, 4, 4)
  for (i in 1:4)
    for (j in 1:4) {
      dd[i, j] <- sqrt (sum (uu (i, j, x, y) * outer (z, z)))
    }
  lbd <- sum (dd * delta) / sum (dd ^ 2)
  xxx[k, ] <- lbd * z
}
lines (xxx[ ,1], xxx[, 2], lwd = 2, col = "BLUE")
abline (0, th0[2] /th0[1], col = "RED", lwd = 2)
th <-sth1$theta
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <- -th
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <-sth2$theta
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <- -th
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <-sth3$theta
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <- -th
points (th[1], th[2], pch = 19, cex = 1.0, col = "BLUE")
th <-sth4$theta
points (th[1], th[2], pch = 19, cex = 1.0, col = "RED")
th <- -th
points (th[1], th[2], pch = 19, cex = 1.0, col = "RED")
th <-sth5$theta
points (th[1], th[2], pch = 19, cex = 1.0, col = "RED")
th <- -th
points (th[1], th[2], pch = 19, cex = 1.0, col = "RED")
```

## Stationary Points

### First Minimum

We zoom in on the first local minimum at (`r sth1$theta`). Its stress is
`r sth1$stress`, and the corresponding configuration has three points in
the corners of an equilateral triangle and the fourth point in its
centroid. Note that this local minimum is a saddle point in
configuration space $\mathbb{R}^{4\times 2}$ (@trosset_mathar_97). The
eigenvalues of $B(\theta)$ are (`r zapsmall(eigen(sth1$b)$values)`) and
those of the Hessian $I-H(\theta)$ are
(`r zapsmall(eigen(sth1$h)$values)`). The area of the contour plot
around the stationary value is in figure \@ref(fig:contfirstmin).

```{r valfirstmin, echo = FALSE, cache = TRUE}
aa <- bb <- seq (.7, 1.2, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

```{r contfirstmin, fig.align= "center", fig.cap = "Contour Plot First Minimum", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
par(pty="s")
contour(aa, bb, z, levels = seq(0.065,.08,length=100))
th <- sth1$theta
points (th[1], th[2], pch = 19, cex = 1.5, col = "BLUE")
```

### Second Minimum

The second local minimum (which is the global minimum) at
(`r zapsmall(sth2$theta)`) has stress `r sth2$stress`. The configuration
are the four points at the corners of a square. The eigenvalues of
$B(\theta)$ are (`r zapsmall(eigen(sth2$b)$values)`) and those of the
Hessian $I-H(\theta)$ are (`r zapsmall(eigen(sth2$h)$values)`). The area
of the contour plot around the stationary value is in figure
\@ref(fig:contsecmin).

```{r values_second_minimum, echo = FALSE, cache = TRUE}
aa <- seq (1, 2, length = 100)
bb <- seq (-.5,.5, length=100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

```{r contsecmin, fig.align= "center", fig.cap = "Contour Plot Second Minimum", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
par(pty="s")
contour(aa, bb, z, levels = seq(0.02,.10,length=100))
th <- sth2$theta
points (th[1], th[2], pch = 19, cex = 1.5, col = "BLUE")
```

### Third Minimum

The third local minimum at (`r sth3$theta`) has stress `r sth3$stress`,
and the corresponding configuration is in figure
\@ref(fig:confthirdmin).

<hr>

```{r confthirdmin, fig.align= "center", fig.cap = "Configuration Third Minimum",echo = FALSE, cache = FALSE}
par(pty="s")
plot(xso3, type = "n", xlab = "dim 1", ylab = "dim 2")
text (xso3, as.character(1:4),col = "RED")
```

The eigenvalues of $B(\theta)$ are (`r zapsmall(eigen(sth3$b)$values)`)
and those of the Hessian $I-H(\theta)$ are
(`r zapsmall(eigen(sth3$h)$values)`). The area of the contour plot
around the stationary value is in figure \@ref(fig:contthirdmin)

```{r values_third_minimum, echo = FALSE, cache = TRUE}
aa <- seq (-.5, .5, length = 100)
bb <- seq (1, 2, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

<hr>

```{r contthirdmin, fig.cap = "Contour Plot Third Minimum",fig.align= "center", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
par(pty="s")
contour(aa, bb, z, levels = seq(.11,.20,length=100))
th <- sth3$theta
points (th[1], th[2], pch = 19, cex = 1.5, col = "BLUE")
```

### First Saddle Point

The saddle point at (`r sth4$theta`) has stress `r sth4$stress`, and the
corresponding configuration is in figure \@ref(fig:conffirstsad).

<hr>

```{r conffirstsad, fig.cap = "Configuration First Saddlepoint", fig.align= "center", echo = FALSE, cache = FALSE}
par(pty="s")
plot(xso4, type = "n", xlab = "dim 1", ylab = "dim 2")
text (xso4, as.character(1:4),col = "RED")
```

The eigenvalues of $B(\theta)$ are (`r zapsmall(eigen(sth4$b)$values)`)
and those of the Hessian $I-H(\theta)$ are
(`r zapsmall(eigen(sth4$h)$values)`). The area of the contour plot
around the stationary value is in figure \@ref(fig:contfirstsad)

```{r values_first_saddlepoint, echo = FALSE, cache = TRUE}
aa <- seq (.30, .34, length = 100)
bb <- seq (1.25, 1.30, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

```{r contfirstsad, fig.cap = "Contour First Saddlepoint", fig.align= "center", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
par(pty="s")
contour(aa, bb, z, levels = seq (.112, .113, length=100))
th <- sth4$theta
points (th[1], th[2], pch = 19, cex = 1.5, col = "RED")
```

### Second Saddle Point

The saddle point at (`r sth5$theta`) has stress `r sth5$stress` and the
corresponding configuration is in figure \@ref(fig:confsecsad)

```{r confsecsad, fig.cap = "Configuration Second Saddlepoint", fig.align= "center", echo = FALSE, cache = FALSE}
par(pty="s")
plot(xso5, type = "n", xlab = "dim 1", ylab = "dim 2")
text (xso5, as.character(1:4),col = "RED")
```

The eigenvalues of $B(\theta)$ are (`r zapsmall(eigen(sth5$b)$values)`)
and those of the Hessian $I-H(\theta)$ are
(`r zapsmall(eigen(sth5$h)$values)`). The area of the contour plot
around the stationary value is in figure \@ref(fig:contsecsad)

```{r values_second_saddlepoint, echo = FALSE, cache = TRUE}
aa <- seq (1.11, 1.14, length = 100)
bb <- seq (.76, .79, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- stress2 (aa[i], bb[j], x, y, delta)
```

```{r contsecsad, fig.cap = "Contour Plot Second Saddlepoint", fig.align= "center", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
par(pty="s")
contour(aa, bb, z, levels = seq (.0672, .0678, length=100))
th <- sth5$theta
points (th[1], th[2], pch = 19, cex = 1.5, col = "RED")
```

## Another Look {#picsphere}

Remember that $\rho(\theta)=\theta'B(\theta)\theta$. Thus
$\sigma(\lambda\theta)=1-\lambda\rho(\theta)+\frac12\lambda^2\theta'\theta$,
and $$
\min_\lambda\sigma(\lambda\theta)=1-\frac12\frac{\rho^2(\theta)}{\theta'\theta}.
$$ Thus we can minimize $\sigma$ over $\theta$ by maximizing $\rho$ over
the unit circle $\mathcal{S}:=\{\theta\mid\theta'\theta=1\}$. This is a
nice formulation, because $\rho$ is norm, i.e. a homogeneous convex
function of $\theta$. Consequently we have transformed the problem from
unconstrained minimization of the DC function (i.e. difference of convex
functions) stress to that of maximization of a ratio of norms. In turn
this is equivalent to maximization of the convex function $\rho$ over
the unit circle, or, again equivalently, over the unit ball, a compact
convex set. This transform was first used in MDS by @deleeuw_C_77,
partly because it made the theory developed by @robert_67 available.

The levels sets $\{\theta\mid\rho(\theta)=\kappa\}$ are the
$\rho$-circles defined by the norm $\rho$. The corresponding
$\rho$-balls $\{\theta\mid\rho(\theta)\leq\kappa\}$ are closed and
nested convex sets containing the origin. Thus we want to find the
largest $\rho$-circle that still intersects $\mathcal{S}$. The
similarity with the geometry of eigenvalue problems is obvious.

```{r ropt, echo = FALSE}
ph<-sth2$theta / sqrt (sum (sth2$theta ^ 2))
ropt <- rho2 (ph[1], ph[2], x, y, delta)
```

In our example we know that the global optimum of stress is at
(`r zapsmall(sth2$theta)`), and if we project that point on the circle
it becomes (`r zapsmall(ph)`). The corresponding optimal $\rho$ is
`r ropt`. Figure \@ref(fig:rhocontour) gives the contourplot for $\rho$,
with the outer $\rho$-circle corresponding with the optimal value. The
fact that the optimal value contour is disjoint from the interior of
$\mathcal{S}$ is necessary and sufficient for global optimality
(@dur_horst_locatelli_98). Notice the sharp corners in the contour plot,
showing the non-diffentiability of $\rho$ at the points where
$d_{12}(\theta)=0$. We could also look for the minimum of $\rho$ on the
unit circle, which means finding the largest $\rho$-circle that touches
$\mathcal{S}$ on the inside. Inspecting figure \@ref(fig:rhocontour)
shows that this will be a point where $\rho$ is not differentiable, i.e.
a point with $d_{12}(\theta)=0$. This minimum $\rho$ problem does not
make much sense in the context of multidimensional scaling, however, and
it not related directly to the minimization of stress.

```{r rho_values, echo = FALSE, cache = TRUE}
aa <- bb <- seq (-1.2, 1.2, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i, j]<- rho2 (aa[i], bb[j], x, y, delta)
```

```{r rhocontour, fig.cap = "Contour Plot for Rho",fig.align= "center", fig.width = 10, fig.height = 10, echo = FALSE, cache = TRUE}
par(pty="s")
contour(aa, bb, z, levels = c(seq(0,1.3,length = 14), ropt), lwd = 2, col = "BLUE")
av <- seq (0,  2 * pi, length = 100)
bv <- sin (av)
cv <- cos (av)
lines(bv,cv,col="RED",lwd=3)
```

## A Final Look {#picline}

Now that we know that the MDS problem is equivalent to maximizing $\rho$
on the unit circle, we can use nonlinear coordinates
$(\theta_1,\theta_2)=(\sin\xi,\cos\xi)$ to reduce the problem to a
one-dimensional unconstrained one in, say, \`\`circle space''. Thus,
with the same abuse of notation as for stress,
$\rho(\xi):=\rho(\sin\xi,\cos\xi)$, and we have to maximize $\rho$ over
$0\leq\xi\leq\pi$.

In figure \@ref(fig:rhononlinearplot) we have plotted $\rho$ as a
function of $\eta$. There are blue vertical lines at the three local
minima in coefficient space, red vertical lines at the stationary
points, and a green vertical line where $d_{12}(\xi)=0$. Note that in
circle space stress has both multiple local minima and multiple local
maxima.

<hr>

```{r rho_nonlinear_values, echo = FALSE, cache = TRUE}
par(pty="s")
av <- seq (0,  pi, length = 10000)
bv <- rep(0, 10000)
sav <- sin (av)
cav <- cos (av)
for (i in 1:10000)
  bv[i] <- rho2 (sav[i], cav[i], x, y, delta)
```

```{r rhononlinearplot, fig.cap = "One-dimensional Rho",fig.align= "center", fig.width = 10, fig.height = 10, echo = FALSE, cache = FALSE}
plot (av, bv, type = "l", col = "RED", lwd = 2, xlab = "xi", ylab = "rho")
th1 <- sth1$theta
th1 <- th1 / sqrt (sum (th1 ^ 2))
xi1 <- asin (th1[1])
abline (v = xi1, col = "BLUE")
th2 <- sth2$theta
th2 <- th2 / sqrt (sum (th2 ^ 2))
xi2 <- asin (th2[1])
abline (v = xi2, col = "BLUE")
th3 <- sth3$theta
th3 <- th3 / sqrt (sum (th3 ^ 2))
xi3 <- asin (th3[1])
abline (v = xi3, col = "BLUE")
th4 <- sth4$theta
th4 <- th4 / sqrt (sum (th4 ^ 2))
xi4 <- asin (th4[1])
abline (v = xi4, col = "RED")
th5 <- sth5$theta
th5 <- th5 / sqrt (sum (th5 ^ 2))
xi5 <- asin (th5[1])
abline (v = xi5, col = "RED")
th0 <- th0 / sqrt (sum (th0 ^ 2))
xi0 <- asin (th0[1])
abline (v = pi+xi0, col = "GREEN")
```

```{r second_derivatives, echo = FALSE, eval = FALSE}
th <- sth1$theta
th <- th / sqrt (sum (th ^ 2))
bh <- bmat2 (th[1], th[2], x, y, delta)
eh <- max (eigen (bh$h)$values)
rh <- rho2 (th[1], th[2], x, y, delta)
hes1 <- eh - rh
th <- sth2$theta
th <- th / sqrt (sum (th ^ 2))
bh <- bmat2 (th[1], th[2], x, y, delta)
eh <- max (eigen (bh$h)$values)
rh <- rho2 (th[1], th[2], x, y, delta)
hes2 <- eh - rh
th <- sth3$theta
th <- th / sqrt (sum (th ^ 2))
bh <- bmat2 (th[1], th[2], x, y, delta)
eh <- max (eigen (bh$h)$values)
rh <- rho2 (th[1], th[2], x, y, delta)
hes3 <- eh - rh
th <- sth4$theta
th <- th / sqrt (sum (th ^ 2))
bh <- 2 (th[1], th[2], x, y, delta)
eh <- max (eigen (bh$h)$values)
rh <- rho2 (th[1], th[2], x, y, delta)
hes4 <- eh - rh
th <- sth5$theta
th <- th / sqrt (sum (th ^ 2))
bh <- bmat2 (th[1], th[2], x, y, delta)
eh <- max (eigen (bh$h)$values)
rh <- rho2 (th[1], th[2], x, y, delta)
hes5 <- eh - rh
```

From lemma xxx we see that the
second derivative $\mathcal{D}^2\rho(\xi)$ is equal to
$\mathbf{tr}\ H(\xi)-\rho(\xi)$. For the three local minima in
coordinate space we find second derivatives
`r c(hes1=0, hes2=0, hes3=0)` in circle space, i.e. they are properly
converted to local maxima. The two stationary points in coordinate space
have second derivatives `r c(hes4=0, hes5=0)`, and are turned into local
minima.

For more general cases, with a basis of $n$ configurations, we know from
@lyusternik-schnirelmann_34 that a continuously differentiable even
function on the unit sphere in $\mathbb{R}^n$ has at least $n$ distinct
pairs of stationary points.

## Discuss

Note that we have used $\sigma$ for three different functions. The first
one with argument $Z$ is defined on *configuration space*, the second
one with argument $\gamma$ on *coefficient space*, and the third one
with argument $\theta$ also on *coefficient space*. This is a slight
abuse of notation, rather innocuous, but we have to keep it in mind.

From lemma xxx we see that
$\mathcal{D}\sigma(X)=\mathcal{D}\sigma(Y)=0$ then
$\mathcal{D}\sigma(1,0)=\mathcal{D}\sigma(0,1)=0$. Thus stationary
points in configuration space are preserved as stationary points in
coefficient space, but the reverse implication may not be true. If
$\mathcal{D}^2\sigma(X)$ and $\mathcal{D}^2\sigma(Y)$ are positive
semi-definite, then so are $\mathcal{D}^2\sigma(1,0)$ and
$\mathcal{D}^2\sigma(0,1)$. Thus local minima are preserved. But it is
entirely possible that $\mathcal{D}^2\sigma(X)$ and/or
$\mathcal{D}^2\sigma(Y)$ are indefinite, and that
$\mathcal{D}^2\sigma(1,0)$ and/or $\mathcal{D}^2\sigma(0,1)$ are
positive semi-definite. Thus saddle points in configuration space can be
mapped into local minima in coefficient space. As we will see this
actually happens with $Y$, the equilateral triangle with center, in our
example.

## Coordinates

```{r coordinates1, echo = FALSE, cache = TRUE}
set.seed(12345)
x <- matrix(rnorm(10), 5, 2)
delta <- dist(x)
eps <- (-500:500) / 100
sy <- rep (0, 1001)
plot (0,
      0,
      xlim = c(-5, 5),
      ylim = c(0, 20),
      xlab = "epsilon",
      ylab = "stress",
      type = "n")
for (i in 1:5) {
  for (j in 1:2) {
    for (k in 1:1001) {
      y <- x
      y[i, j] <- x[i, j] + eps[k]
      dy <- dist (y)
      sy[k] <- sum ((delta - dy) ^ 2)
    }
    lines (eps, sy, lwd = 2, col = "RED")
  }
}
```

```{r coordinates2, echo = FALSE, cache = TRUE}
set.seed(12345)
x <- matrix(rnorm(10), 5, 2)
delta <- dist(matrix(rnorm(10),5,2))
eps <- (-500:500) / 100
sy <- rep (0, 1001)
plot (0,
      0,
      xlim = c(-5, 5),
      ylim = c(0, 20),
      xlab = "epsilon",
      ylab = "stress",
      type = "n")
for (i in 1:5) {
  for (j in 1:2) {
    for (k in 1:1001) {
      y <- x
      y[i, j] <- x[i, j] + eps[k]
      dy <- dist (y)
      sy[k] <- sum ((delta - dy) ^ 2)
    }
    lines (eps, sy, lwd = 2, col = "RED")
  }
}
```

```{r sq4, echo = FALSE}
a <- (2 + sqrt (2)) / 8
xtr <- matrix (c(a,a,-a,-a,a,-a,a,-a), 4, 2)
d <- dist(xtr)
delta <- dist(diag(4))
delta <- delta/sqrt(sum(delta^2))
lbd <- sum (delta * d) / sum (d ^ 2)
xtr <- xtr * lbd
d <- d * lbd
str <- sum((delta-d)^2)
b <- -as.matrix (delta / d)
diag(b) <- -rowSums(b)
g<-4 * xtr - b%*% xtr
ev<-eigen (b / 4)$values
```

```{r tr4, echo = FALSE, cache = TRUE}
a <- (1 + sqrt (3)) / 4
b <- (3 + sqrt (3)) / 8
xsq <- matrix (c(a,-a/2,-a/2,0,0,-b,b,0), 4, 2)
d <- dist (xsq)
delta <- dist(diag(4))
delta <- delta/sqrt(sum(delta^2))
lbd <- sum (delta * d) / sum (d ^ 2)
xsq <- xsq * lbd
d <- d * lbd
ssq <-sum((delta-d)^2)
b <- -as.matrix (delta / d)
diag(b) <- -rowSums(b)
g<-4 * xsq - b%*% xsq
e<-eigen (b / 4)$values
```

Let's look at a small example with four points, all dissimilarities
equal, and all weights equal to one. There is a local minimum with four
points in the corners of a square, with stress equal to `r str`. And
there is another local minimum with three points forming an equilateral
triangle, and the fourth point in the center. This has stress `r ssq`.
We can compute stress for all points of the form $\alpha X+\beta Y$,
where $X$ and $Y$ are the two local minima. Figure
\@ref(fig:equalcontourplot) has a contour plot of
$\sigma(\alpha,\beta)$, showing the local minima at $(1,0)$ and $(0,1)$,
and the local maximum at $(0,0)$.

```{r qcont, echo = FALSE, cache = TRUE}
stresser <- function (a, b) {
  u <- a * xsq + b * xtr
  d <- dist (u)
  return (sum((delta - d) ^ 2))
}
xl <- yl <- seq (-.5, 1.5, length = 100)
z <- matrix (0, 100, 100)
for (i in 1:100) for (j in 1:100)
  z[i,j] <- stresser (xl[i], yl[j])
za <- rep(0, 1000)
xa <- seq (-.5, 1.5, length = 1000)
for (i in 1:1000)
  za[i] <- stresser(xa[i], 1 - xa[i])
```

```{r equalcontourplot, fig.align = "center", echo = FALSE, fig.cap = "Plane spanned by two local minima, equal dissimilarities"}
contour (xl, yl, z, levels = seq(0, 1, length = 100), col = "RED")
points (rbind (diag (2), c(0,0)), pch = 19)
```

Alternatively, we can plot stress on the line connecting $X$ and $Y$.
Note that although stress only has a local maximum at the origin in
configuration space, it can very well have local maxima if restricted to
lines. In fact on a line connecting two local minima there has to be at
least one local maximum.

```{r equallineplot, fig.align = "center", echo = FALSE, fig.cap="Line connecting two local minima, equal dissimilarities"}
plot (xa, za, type = "l", col = "RED", ylab = "stress", xlab = "alpha", lwd =2)
abline(v=0)
abline(v=1)
```
