# Splinical MDS {#chsplinical}

## Splines {#mathsplines}

In this section we give a short introduction, with examples, to (univariate) splines, B-splines, and I-splines. It is taken from @deleeuw_E_17i, with some edits to make it fit into the book. The report it was taken from has more detail and more examples.

To define *spline functions*  we first define a finite sequence of *knots* $T=\{t_j\}$ on the real line, with $t_1\leq\cdots\leq t_p,$ and an *order* $m$. In addition each knot $t_j$ has a *multiplicity* $m_j$, the number of knots equal to $t_j$. We suppose throughout that $m_j\leq m$ for all $j$.

A function $f$ is a *spline function of order* $m$ for a knot sequence $\{t_j\}$ if

1. $f$ is a polynomial $\pi_j$ of degree at most $m-1$ on each half-open interval $I_j=[t_j,t_{j+1})$ for $j=1,\cdots,p$, 
2. the polynomial pieces are joined in such a way that $\mathcal{D}^{(s)}_-f(t_j)=\mathcal{D}^{(s)}_+f(t_j)$ for $s=0,1,\cdots,m-m_j-1$ and $j=1,2,\cdots,p$.

Here we use $\mathcal{D}^{(s)}_-$ and $\mathcal{D}^{(s)}_+$ for the left and right $s^{th}$-derivative operator. If $m_j=m$ for some $j$, then the second requirement is empty, if $m_j=m-1$ then the second requirement means $\pi_j(t_j)=\pi_{j+1}(t_j)$, i.e. we require continuity of $f$ at $t_j$. If $1\leq m_j<m-1$ then $f$ must be $m-m_j-1$ times differentiable, and thus continuously differentiable, at $t_j$. 

In the case of simple knots (with multiplicity one) a spline function of order one is a *step function* which steps from one level to the next at each knot. A spline of order two is piecewise linear, with the pieces joined at the knots so that the spline function is continuous. Order three means a piecewise quadratic function which is continuously differentiable at the knots. And so on. 

### B-splines {#mathbsplines}

Alternatively, a spline function of order $m$ can be defined as a linear combination of *B-splines* (or *basic splines*) of order $m$ on the same knot sequence. A B-spline of order $m$ is a spline function consisting of at most $m$ non-zero polynomial pieces. A B-spline $\mathcal{B}_{j,m}$ is determined by the $m+1$ knots $t_j\leq\cdots\leq t_{j+m}$, is zero outside the interval $[t_j,t_{j+m})$, and positive in the interior of that interval. Thus if $t_j=t_{j+m}$ then $\mathcal{B}_{j,m}$ is identically zero.

For an arbitrary finite knot sequence $t_1,\cdots,t_p$, there are $p-m$ B-splines to of order $m$ to be considered, although some may be identically zero. Each of the splines covers at most $m$ consecutive intervals, and at most $m-1$ different B-splines are non-zero at each point. 

#### Boundaries {#mathbboundaries}

B-splines are most naturally and simply defined for doubly infinite sequences of knots, that go to $\pm\infty$ in both directions. In that case we do not have to worry about boundary effects, and each subsequence of $m+1$ knots defines a B-spline. For splines on finite sequences of $p$ knots we have to decide what happens at the boundary points. 

There are B-splines for $t_j,\cdots,t_{j+m}$ for all $j=1,\cdots,p-m$. This means that the first $m-1$ and the last
$m-1$ intervals have fewer than $m$ splines defined on them. They are not part of what @deboor_01, page 94, calls the *basic interval*. For doubly infinite sequences of knots there is not need to consider such a basic interval.

If we had $m$ additional knots on both sides of our knot sequence we would also have $m$ additional B-splines for $j=1-m,\cdots,0$ and $m$ additional B-splines for $j=p-m+1,\cdots,p$. By adding these additional knots we make sure each interval $[t_j,t_{j+1})$ for $j=1,\cdots,p-1$ has $m$ B-splines associated with it. There is stil some ambiguity on what to do at $t_p$, but we can decide to set the value of the spline there equal to the limit from the left, thus making the B-spline left-continuous there.

In our software we will use the convention to define our splines on a closed interval $[a,b]$ with $r$ *interior knots*
$a<t_1<\cdots<t_r<b$, where interior knot $t_j$ has multiplicity $m_j$. We extend this to a series of $p=M+2m$ knots, with $M=\sum_{j=1}^r m_j$, by starting with $m$ copies of $a$, appending $m_j$ copies of $t_j$ for each $j=1,\cdots,r$, and finishing with $m$ copies of $b$. Thus $a$ and $b$ are both knots with multiplicity $m$. This defines the *extended partition* (@schumaker_07, p 116), which is just handled as any knot sequence would normally be. 

#### Normalization {#mathbnormalize}

The conditions we have mentioned only determine the B-spline up to a normalization. There are two popular ways of normalizing B-splines. The $N$-splines $N_{j,m}$, a.k.a. the *normalized B-splines* $j$ or order $m$, satisfies
\begin{equation}\label{E:nsum}
\sum_{j}N_{j,m}(t)=1.
\end{equation}
Note that in general this is not true for all $t$, but only for all $t$ in the *basic interval*. 

Alternatively we can normalize to $M$-splines, for which
\begin{equation}\label{E:mint}
\int_{-\infty}^{+\infty}M_{j,m}(t)dt=\int_{t_j}^{t_{j+k}}M_{j,m}(t)dt=1.
\end{equation}
There is the simple relationship
\begin{equation}\label{E:NM}
N_{j,m}(t)=\frac{t_{j+m}-t_j}{m}\ M_{j,m}(t).
\end{equation}

#### Recursion {#mathbrecursion}

B-splines can be defined in various ways, using piecewise polynomials, divided differences, or recursion. The recursive definition, first used as the preferred definition of B-splines by @deboor_hollig_85, is the most convenient one for computational purposes, and that is the one we use. 

The recursion is due independently to @cox_72 for simple knots and to @deboor_72 in the general case, is
\begin{equation}\label{E:Mspline}
M_{j,m}(t)=\frac{t-t_j}{t_{j+m}-t_j}M_{j,m-1}(t)+\frac{t_{j+m}-t}{t_{j+m}-t_j}M_{j+1,m-1}(t),
\end{equation}
or
\begin{equation}\label{E:Nspline}
N_{j,m}(t)=\frac{t-t_j}{t_{m+j-1}-t_j}N_{j,m-1}(t)+\frac{t_{j+m}-t}{t_{j+m}-t_{j+1}}N_{j+1,m-1}(t).
\end{equation}

A basic result in the theory of B-splines is that the different B-splines are linearly independent and form a basis for the linear space of spline functions (of a given order and knot sequence). 

In section \@ref(apcodemathadd) the basic BSPLVB algorithm from @deboor_01, page 111, for normalized B-splines is translated to R and C. There are two auxiliary routines, one to create the extended partition, and one that uses bisection to locate the knot interval in which a particular value is located (@schumaker_07, p 191). The R function bsplineBasis() takes an arbitrary knot sequence. It can be combined with extendPartition(), which uses inner knots and boundary points to create the extended partion.

#### Illustrations

For our example, which is the same as the one from figure 1 in @ramsay_88, we choose $a=0$, $b=1$, with simple interior knots `r c(.3,.5,.6)`. First the *step functions*, which have order 1. 

```{r order1mult1, fig.align = "center", fig.cap = "Zero Degree Splines with Simple Knots", echo = FALSE}
innerknots <- c(.3, .5, .6)
multiplicities <- c(1, 1, 1)
order <- 1
lowend <- 0
highend <- 1
x <- seq (1e-6, 1 - 1e-6, length = 1000)
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
k <- ncol (h)
par (mfrow = c(2, 2))
for (j in 1:k) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab,
    ylim = c(0, 1)
  )
}
```

Now the *hat functions*, which have order 2, again with simple knots.

```{r order2mult1, fig.align = "center", fig.cap = "Piecewise Linear Splines with Simple Knots", echo = FALSE}
multiplicities <- c(1, 1, 1)
order <- 2
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
k <- ncol (h)
par (mfrow = c(2, 3))
for (j in 1:k) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab,
    ylim = c(0, 1)
  )
}
```

Next piecewise quadratics, with simple knots, which implies continuous differentiability at the knots. This are the N-splines corresponding with the M-splines in figure 1 of @ramsay_88.

```{r order3mult1, fig.align = "center", fig.cap = "Piecwise Quadratic Splines with Simple Knots", echo = FALSE}
multiplicities <- c(1, 1, 1)
order <- 3
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
k <- ncol (h)
par (mfrow = c(2, 3))
for (j in 1:k) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab,
    ylim = c(0, 1)
  )
}
```

If we change the multiplicities to `r c(1,2,3)`, then we lose some of the smoothness.

```{r order3mult2, fig.align = "center", fig.cap = "Piecewise Quadratic Splines with Multiple Knots", echo = FALSE}
multiplicities <- c(1, 2, 3)
order <- 3
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
k <- ncol (h)
par (mfrow = c(3, 3))
for (j in 1:k) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab,
    ylim = c(0, 1)
  )
}
```

### I-splines {#mathisplines}

There are several ways to require splines to be monotone increasing. Since B-splines are non-negative, the definite integral of a B-spline of order $m$ from the beginning of the interval to a value $x$ in the interval is an increasing spline of order $m+1$. Integrated B-splines are known as *I-splines* (@ramsay_88). Non-negative linear combinations I-splines can be used as a basis for the convex cone of increasing splines. Note, however, that if we use an extended partition, then all I-splines start at value zero and end at value one, which means their convex combinations are those
splines that are also probability distributions on the interval. To get a basis for the increasing splines we need to
add the constant function to the I-splines and allow it to enter the linear combination with either sign.

I-splines are most economically computed by using the formula first given by @gaffney_76. If $\ell$ is defined by $t_{j+\ell-1}\leq x<t_{j+\ell}$ then
$$
\int_{x_j}^x M_{j,m}(t)dt=\frac{1}{m}\sum_{r=0}^{ \ell-1}(x-x_{j+r})M_{j+r,m-r}(x)
$$
It is somewhat simpler, however, to use lemma 2.1 of @deboor_lyche_schumaker_76. This says
$$
\int_a^xM_{j,m}(t)dt=\sum_{\ell\geq j}N_{\ell,m+1}(x)-\sum_{\ell\geq j}N_{\ell,m+1}(a),
$$
If we specialize this to I-splines, we find , as in @deboor_76, formula 4.11,
$$
\int_{-\infty}^x M_{j,m}(t)dt=\sum_{\ell=j}^{j+r}N_{\ell,m+1}(x)
$$
for $x\leq t_{j+r+1}$. This shows that I-splines can be computed by using cumulative sums of B-spline values.

Note that using the definition using integration does not give a natural way to define increasing splines of degree one, i.e. increasing step functions. There is no such problem with the cumulative sum approach.

#### Increasing Coefficients

As we know, a spline is a linear combination of B-splines. The formula for the derivative of a spline, for example in @deboor_01, p 116, shows that a spline is increasing if the coefficients of the linear combination of B-splines are increasing. Thus we can fit an increasing spline by restricting the coefficients of the linear combination to be increasing, again using the B-spline basis. 

It turns out this is in fact identical to using I-splines. If the B-spline values at $n$ points are in an $n\times r$ matrix $H$, then non-decreasing coefficients $\beta$ are of the
form $\beta=S\alpha+\gamma e_r$, where $S$ is lower-diagonal with all elements on and below the diagonal equal to one, where $\alpha\geq 0$, where $e_r$ has all elements
equal to one, and where $\gamma$ can be of any sign. So $H\beta=(HS)\alpha+\gamma e_n$. Thus non-decreasing coefficients is the same thing as using cumnulative sums of the B-spline basis.

#### Increasing Values

Finally, we can simply require that the $n$ elements of $H\beta$ are increasing. This is a less restrictive requirement, because it allows for the possibility that the spline is decreasing between data values. It has the rather serious disadvantage, however, that it does
its computations in $n$-dimensional space, and not in $r$-dimensional space, where $r=M+m$, which is usually much smaller than $n$. Software for the increasing-value restrictions has been
written by @deleeuw_E_15e. In our software, however, we prefer the `cumsum()` approach. It is less general, but considerably more efficient.

We use the same Ramsay example as before, but now cumulatively. First we integrate step functions with simple knots, which have order one, using `isplineBasis()`. The corresponding I-splines are piecewise linear with order two.

```{r Iorder1mult1, echo = FALSE, fig.align = "center", fig.cap = "Not Sure"}
multiplicities <- c(1,1,1)
order <- 2
knots <- extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- isplineBasis (x, knots, order)
par (mfrow=c(2,2))
for (j in 2:ncol(h)) { 
  ylab <- paste("B-spline", formatC(j, digits = 1, width = 2, format = "d"))
  plot (x, h[, j], type="l", col = "RED", lwd = 3, ylab = ylab)
}
```

Now we integrate the hat functions, which have order 2, again with simple knots, to find piecewise quadratic I-splines of order 3. These are the functions in the example of @ramsay_88.

```{r Iorder2mult1, echo = FALSE, fig.align = "center", fig.cap = "Monotone Piecewise Linear Splines with Simple Knots"}
multiplicities <- c(1, 1, 1)
order <- 3
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- isplineBasis (x, knots, order)
par (mfrow = c(2, 3))
for (j in 2:ncol(h)) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab
  )
}
```

Finally, we change the multiplicities to `r c(1,2,3)`, and compute the corresponding piecewise quadratic I-splines.

```{r Iorder3mult2, echo = FALSE, fig.align = "center", fig.cap = "Monotone Piecewise Quadratic Splines with Multiple Knots"}
multiplicities <- c(1, 2, 3)
order <- 3
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- isplineBasis (x, knots, order)
par (mfrow = c(3, 3))
for (j in 2:ncol(h)) {
  ylab <-
    paste("B-spline", formatC(
      j,
      digits = 1,
      width = 2,
      format = "d"
    ))
  plot (
    x,
    h[, j],
    type = "l",
    col = "RED",
    lwd = 3,
    ylab = ylab
  )
}
```

### Time Series Example

Our first example smoothes a time series by fitting a spline. We use the number of births in New York from 1946 to 1959 (on an unknown scale), from Rob Hyndman's time series archive.


#### B-splines

First we fit B-splines of order three. The basis matrix uses $x$ equal to $1:168$,
with inner knots `r 12 * 1:13`, and interval $[1,168]$. 
```{r births_basis}
innerknots <- 12 * 1:13
multiplicities <- rep(1, 13)
lowend <- 1
highend <- 168
order <- 3
x <- 1:168
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
u <- lm.fit(h, births)
res <- sum ((births - h %*% u$coefficients) ^ 2) / 2
```

```{r birthbsplineplot, fig.align = "center", echo = FALSE, cache = TRUE, fig.cap = "Monotone Piecewise Quadratic Splines with Simple Knots"}
birthsts <- ts (cbind (births, h%*%u$coefficients), frequency = 12, start = c(1946, 1))
plot (birthsts, lwd = 2, plot.type = "single", col = c("RED", "BLUE"))
```

The residual sum of squares is `r res`.

#### I-splines

We now fit the I-spline using the B-spline basis. Compute $Z=HS$ using `cumsum()`, and then $\overline y$ and $\overline Z$ by centering (substracting the column means). The formula is
$$
\min_{\alpha\geq 0,\gamma}\mathbf{SSQ}\ (y-Z\alpha-\gamma e_n)=\min_{\alpha\geq 0}\mathbf{SSQ}\ (\overline y-\overline Z\alpha).
$$
We use `pnnls()` from @wang_lawson_hanson_15.

```{r births_isplines_comp, cache = TRUE}
knots <- extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- isplineBasis (x, knots, order)
g <- cbind (1, h[,-1])
u <- pnnls (g, births, 1)$x
v <- g%*%u
```

```{r birthsisplinesplot, fig.align = "center", echo = FALSE, fig.cap = "Monotone Piecewise Linear Splines with Simple Knots"}
birthsts <- ts (cbind (births, v), frequency = 12, start = c(1946, 1))
plot (birthsts, lwd = 2, plot.type = "single", col = c("RED", "BLUE"))
```

The residual sum of squares is `r sum ((births - v)^2) / 2`.

#### B-Splines with monotone weights

Just to make sure, we also solve the problem
$$
\min_{\beta_1\leq\beta_2\leq\cdots\leq\beta_p}\mathbf{SSQ}(y-X\beta),
$$
which should give the same solution, and the same loss function value, because it is just another way to fit I-splines. We use the `lsi()` function from @wang_lawson_hanson_15.

```{r births_isplines_values, cache = TRUE}
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
nb <- ncol (h)
d <- matrix(0, nb - 1, nb)
diag(d) = -1
d[outer(1:(nb - 1), 1:nb, function(i, j)
  (j - i) == 1)] <- 1
u <- lsi(h, births, e = d, f = rep(0, nb - 1))
v <- h %*% u
```

```{r birthsisplinesvaluesplot, fig.align = "center", echo = FALSE, fg.cap = "Monotone Piecewise Quadratic Splines with Simple Knots"}
birthsts <- ts (cbind (births, v), frequency = 12, start = c(1946, 1))
plot (birthsts, lwd = 2, plot.type = "single", col = c("RED", "BLUE"))
```

The residual sum of squares is `r sum ((births - v)^2) / 2`, 
indeed the same as before.

#### B-Splines with monotone values

Finally we solve

$$
\min_{x_1'\beta\leq\cdots\leq x_n'\beta} \mathbf{SSQ}\ (y-X\beta)
$$

using qpmaj() from section ???.

```{r births_bsplines_monotonic_comp, cache = TRUE}
knots <-
  extendPartition (innerknots, multiplicities, order, lowend, highend)$knots
h <- bsplineBasis (x, knots, order)
a <- diff(diag(nrow(h))) %*% h
u <- qpmaj(births, h = h, a = a)
```

```{r birthsbsplinesmonotonicplot,fig.align = "center", echo = FALSE, fig.cap = "Monotone Piecewise Quadratic Splines with Multiple Knots"}
birthsts <- ts (cbind (births, u$predict), frequency = 12, start = c(1946, 1))
plot (birthsts, lwd = 2, plot.type = "single", col = c("RED", "BLUE"))
```

The residual sum of squares is `r u$ftotal` , which is indeed smaller than the I-splines value, although only very slightly so.

### Local positivity, monotonicity, convexity

