# (APPENDIX) Appendices {.unnumbered}

# Code {#apcode}

## R Code {#rapcode}

The MDS functions in R throughout work with square matrices of
weights, dissimilarities, and distances. More efficient versions, 
that have their computations done in C, will be added along 
the way.

### utilities.R {#aputilitiescode}

```{r file_auxilary, code = readLines("rcode/utilities.R"), eval = FALSE}
```

### common/indexing.r {#apindexingcode}

```{r file_auxilary, code = readLines("rcode/common/indexing.R"), eval = FALSE}
```

### common/io.r {#apiocode}

```{r file_auxilary, code = readLines("rcode/common/io.R"), eval = FALSE}
```

### common/linear.r {#aplinearcode}

```{r file_auxilary, code = readLines("rcode/common/linear.R"), eval = FALSE}
```

### common/nextPC.r {#appermcode}

```{r file_auxilary, code = readLines("rcode/common/nextPC.R"), eval = FALSE}
```

### common/smacof.r {#apsmacofcode}

```{r file_auxilary, code = readLines("rcode/common/smacof.R"), eval = FALSE}
```

### properties.R {#apcodeprop}

```{r file_auxilary, code = readLines("rcode/properties.R"), eval = FALSE}
```

### expandOneDim.R {#apcodeexpand}

```{r file_auxilary, code = readLines("rcode/expandOneDim.R"), eval = FALSE}
```

### pictures.R {#apcodepic}

```{r file_auxilary, code = readLines("rcode/pictures.R"), eval = FALSE}
```

### classical.R {#apcodeclass}

```{r file_auxilary, code = readLines("rcode/classical.R"), eval = FALSE}
```

### minimization.R {#apcodeminim}

```{r file_auxilary, code = readLines("rcode/minimization.R"), eval = FALSE}
```

### full.R {#apcodefull}

```{r file_auxilary, code = readLines("rcode/full.R"), eval = FALSE}
```

### unfolding.R {#apcodeunfold}

```{r file_auxilary, code = readLines("rcode/unfolding.R"), eval = FALSE}
```

### constrained.R {#apcodencons}

```{r file_auxilary, code = readLines("rcode/constrained.R"), eval = FALSE}
```

### nominal.R {#apcodenominal}

```{r file_auxilary, code = readLines("rcode/nominal.R"), eval = FALSE}
```

### sstress.R {#apcodesstress}

```{r file_auxilary, code = readLines("rcode/sstress.R"), eval = FALSE}
```

### inverse.R {#apcodeinverse}

```{r file_auxilary, code = readLines("rcode/inverse.R"), eval = FALSE}
```

### global.R {#apcodeglobal}

```{r file_auxilary, code = readLines("rcode/global.R"), eval = FALSE}
```

### mathadd.R {#apcodemathadd}

```{r file_auxilary, code = readLines("rcode/mathadd.R"), eval = FALSE}
```

## C code

### deboor.c

```{c file_auxilary, code = readLines("ccode/deboor.c"), eval = FALSE}
```

### cleanup.c

```{c file_auxilary, code = readLines("ccode/cleanup.c"), eval = FALSE}
```

### jacobi.c

```{c file_auxilary, code = readLines("ccode/jacobi.c"), eval = FALSE}
```

### jbkTies.c

```{c file_auxilary, code = readLines("ccode/jbkTies.c"), eval = FALSE}
```

### matrix.c

```{c file_auxilary, code = readLines("ccode/matrix.c"), eval = FALSE}
```

### jeffrey.c

```{c file_auxilary, code = readLines("ccode/jeffrey.c"), eval = FALSE}
```

### smacofBlockSort.c

```{c file_auxilary, code = readLines("ccode/smacofBlockSort.c"), eval = FALSE}
```

### smacofConvert.c

```{c file_auxilary, code = readLines("ccode/smacofConvert.c"), eval = FALSE}
```

### nextPC.c

```{c file_auxilary, code = readLines("ccode/nextPC.c"), eval = FALSE}
```

# Data {#apdatasets}

## Small {#apdatadsmall}

```{r dsmall, echo = FALSE}
dsmall
```

## De Gruijter {#apdatagruijter}

```{r gruijter, echo = FALSE}
gruijter
```

## Ekman {#apdataekman}

```{r ekman_back, echo = FALSE}
#ekman
```

## Vegetables {#apdataveg}

```{r}
veg <- abs (qnorm (matrix (c(.500,.818,.770,.811,.878,.892,.899,.892,.926,
.182,.500,.601,.723,.743,.736,.811,.845,.858,
.230,.399,.500,.561,.736,.676,.845,.797,.818,
.189,.277,.439,.500,.561,.588,.676,.601,.730,
.122,.257,.264,.439,.500,.493,.574,.709,.764,
.108,.264,.324,.412,.507,.500,.628,.682,.628,
.101,.189,.155,.324,.426,.372,.500,.527,.642,
.108,.155,.203,.399,.291,.318,.473,.500,.628,
.074,.142,.182,.270,.236,.372,.358,.372,.500), 9, 9)))
```

# Unlicense {#apunlicense}

This book is free and unencumbered text and software released into the public domain.

Anyone is free to copy, modify, publish, use, compile, sell, or
distribute this book, either in source code form or as a compiled
binary, for any purpose, commercial or non-commercial, and by any
means.

In jurisdictions that recognize copyright laws, the author 
of this book dedicates any and all copyright interest in the
software to the public domain. I make this dedication for the benefit
of the public at large and to the detriment of my heirs and
successors. I intend this dedication to be an overt act of
relinquishment in perpetuity of all present and future rights to this
book under copyright law.

THE BOOK IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
IN NO EVENT SHALL THE AUTHOR BE LIABLE FOR ANY CLAIM, DAMAGES OR
OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
ARISING FROM, OUT OF OR IN CONNECTION WITH THE BOOK OR THE USE OR
OTHER DEALINGS IN THE BOOK.

For more information, please refer to <http://unlicense.org/>

# References {.unnumbered}

