---
title: "Exceedingly Simple Sorting with Indices"
author: "Jan de Leeuw"
date: "Version 01, March 22, 2017"
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
graphics: yes
fontsize: 12pt
abstract: We use the system `qsort` to write a routine that produces both the sort
  an the order of a vector of doubles.
---
<style type="text/css">

body{ /* Normal  */
   font-size: 18px;
}
td {  /* Table  */
   font-size: 18px;
}
h1 { /* Header 1 */
 font-size: 28px;
 color: DarkBlue;
}
h2 { /* Header 2 */
 font-size: 22px;
 color: DarkBlue;
}
h3 { /* Header 3 */
 font-size: 18px;
 color: DarkBlue;
}
code.r{ /* Code block */
  font-size: 18px;
}
pre { /* Code block */
  font-size: 18px
}
</style>


```{r function_code, echo = FALSE}
source("mySort.R")
```
```{r packages, echo = FALSE}
options (digits = 10) 
library (microbenchmark)
```
Note: This is a working paper which will be expanded/updated frequently. All suggestions for improvement are welcome. The directory [deleeuwpdx.net/pubfolders/mysort](http://deleeuwpdx.net/pubfolders/mysort) has a pdf version, the complete Rmd file with the code chunks, 
and the R source code.

#Introduction

This may not be of much use to anyone else, but I'll make it available anyway. For a larger project I needed something simple in C to sort a vector and return both
the sorted vector and the corresponding index vector. Thus it combines `sort` and `order` in R. It creates an array of structures
in C, where each structure has a value member and an index member. It then sorts the structures using the system `qsort`, by comparison
of the values. Finally it splits the sorted array of structures into values and indices and returns those to R in a two-element list.

#Example

```{r example}
set.seed (12345)
x <- rnorm (10)
mySort (x)
```
as compared to

```{r standard}
list (values = sort (x), indices = order (x))
```

A quick comparison shows very little timing difference.

```{r compare}
f <- function () {
  x <- rnorm(10000)
  h <- mySort (x)
}
g <- function () {
  x <- rnorm (10000)
  h <- list (values = sort (x), indices = order (x))
}
microbenchmark (f, g, times = 1000)
```


#C code

```{c c_code, engine='c', results='hide'}
#include <math.h>
#include <stdlib.h>

int myComp(const void *, const void *);
void mySort(double *, int *, int *);

struct couple {
    double value;
    int index;
};

int myComp(const void *px, const void *py) {
    double x = ((struct couple *)px)->value;
    double y = ((struct couple *)py)->value;
    return (int)copysign(1.0, x - y);
}

void mySort(double *x, int *k, int *n) {
    int nn = *n;
    struct couple *xi =
        (struct couple *)calloc((size_t)nn, (size_t)sizeof(struct couple));
    for (int i = 0; i < nn; i++) {
        xi[i].value = x[i];
        xi[i].index = i + 1;
    }
    (void)qsort(xi, (size_t)nn, (size_t)sizeof(struct couple), myComp);
    for (int i = 0; i < nn; i++) {
        x[i] = xi[i].value;
        k[i] = xi[i].index;
    }
    free(xi);
}
```

#R code

```{r r_code}
dyn.load("mySort.so")

mySort <- function (x) {
  h <-
    .C(
      "mySort",
      values = as.double (x),
      indices = as.integer(1:length(x)),
      as.integer(length(x))
    )
  return (list (values = h$values, indices = h$indices))
}
```