---
title: "Partial inverses by Sweeping"
author: "Jan de Leeuw"
date: "March 21, 2016"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
graphics: yes
bibliography: sweep.bib
---

```{r exec_code, echo = FALSE}

dyn.load("hsweep.so")

rSweep <- function (a, indi) {
for (j in indi) {
    pv <- a[j, j]
    if (pv == 0) next ()
    pr <- a[j, -j]
    pc <- a[-j, j]
    a[j, j] <- -1 / pv
    a[j, -j] <- pr / pv
    a[-j, j] <- pc / pv
    a[-j, -j] <- a[-j, -j] - outer(pc, pr) / pv
    }
return (a)
}

fSweep <- function (b , ind, eps = 1e-10, trianin = FALSE, trianout = FALSE) {
	if (trianin) {
		m <- length (b)
		n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
		a <- b
	} else {
		n <- nrow (b)
		m <- (n * n + n) / 2
		a <- b[upper.tri (b, diag = TRUE)]
	}
	for (k in ind) {
		a <- .Fortran("gsweep",
			diago = as.double (a[cumsum (1:n)]),
			trian = as.double (a),
			k = as.integer (k),
			l = as.integer (0),
			m = as.integer (m),
			n = as.integer (n),
			e = as.double (eps),
			ifault = as.integer (0))$trian
		}
	if (trianout) {
		return (a)
	} else {
		return (tri2sym (a))
	}
}

cSweep <- function (b , ind, eps = 1e-10, trianin = FALSE, trianout = FALSE) {
	if (trianin) {
		m <- length (b)
		n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
		a <- b
	} else {
		n <- nrow (b)
		m <- (n * n + n) / 2
		a <- b[upper.tri (b, diag = TRUE)]
	}
	a <- .C("hsweep",
		diago = as.double (a[cumsum (1:n)]),
		trian = as.double (a),
		n = as.integer (n),
		m = as.integer (m),
		ind = as.integer (ind),
		npiv = as.integer (length (ind)),
		eps = as.double (eps))$trian
	if (trianout) {
		return (a)
	} else {
		return (tri2sym (a))
	}
}

tri2sym <- function (a) {
	m <- length (a)
	n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
	b <- matrix (0, n, n)
	b[upper.tri (b, diag = TRUE)] <- a
	b <- b + t(b)
	diag (b) <- diag (b) / 2
	return (b)
}
```

#Simple Sweep

Suppose \(A\) is a square non-singular matrix. A *simple sweep* on the \(k^{th}\) element replaces \(A\) by the matrix with elements

\begin{align*}
a_{kk}&=\frac{1}{a_{kk}},\\
a_{ki}&=-\frac{a_{ki}}{a_{kk}},\\
a_{ik}&=\frac{a_{ik}}{a_{kk}},\\
a_{ij}&=a_{ij}-\frac{a_{ik}a_{kj}}{a_{kk}}
\end{align*}
for \(j\not= k\) and \(i\not= k\).

In R that becomes
```{r sweep_r}
sweep <- function (a, k) {
  p <- a [k, k] 
  a [k, k] <- 1 /p
  a [-k, -k] <- a [-k, -k] - outer (a [-k, k], a [k, -k]) / p
  a [k, -k] <- -a[k, -k] / p
  a [-k, k] <- a[-k, k] / p
  return (a)
}
```
#Example Sweep

```{r mprint, echo = FALSE}
mprint <- function (x, d = 2, w = 5) {
  print (noquote (formatC (x, di = d, wi = w, fo = "f")))
}
```
```{r}
set.seed(12345)
mprint (a <- matrix (sample(1:9), 3, 3))
mprint (sweep(a, 1))
```
#Using Sweeping for the Inverse

```{r}
sweepSubset <- function (a, k, verbose = FALSE) {
  for (i in k) {
    a <- sweep (a, i)
    if (verbose) mprint (a)
  }
  return (a)
}
```
```{r}
mprint (b <- cSweep (a, 1 : nrow (a)))
mprint (b <- fSweep (a, 1 : nrow (a)))
mprint (b <- rSweep (a, 1 : nrow (a)))
mprint (a %*% b)
```
Using Sweeping for Regression
-------
```{r}
x <- matrix (rnorm (20), 10, 2)
y <- rnorm (10)
mprint (b<-qr.solve (x, y))
mprint (sum((y-x%*%b) ^ 2))
h <- crossprod (cbind (x,y))
mprint (sweepSubset (h, 1 : 2, verbose = TRUE))
```
The partial inverse
----
Suppose \(k\) is a subset of \(\{1,2,\cdots,n\}\) and \(\overline{k}\) is the complimentary subset. In our examples we usually take, for convenience, \(k=\{1,2,\cdots,k\}\). The *partial inverse* of \(A\) on \(k\), which generalizes the simple sweep, is
\[
\mathbf{inv}_k(A)=\begin{bmatrix}
A_{kk}^{-1}&-A_{kk}^{-1}A_{k\overline{k}}\\
A_{\overline k k}A_{kk}^{-1}&A_{\overline k\overline k}-A_{\overline k k}A_{kk}^{-1}A_{k\overline{k}}
\end{bmatrix}.
\]

Why is this called a partial inverse ? If
\[
\begin{bmatrix}
A_{kk}&A_{k\overline k}\\
A_{\overline k k}&A_{\overline k\overline k}
\end{bmatrix}
\begin{bmatrix}
x\\y
\end{bmatrix}=
\begin{bmatrix}
u\\v
\end{bmatrix},
\]
then
\[
\begin{bmatrix}
A_{kk}^{-1}&-A_{kk}^{-1}A_{k\overline{k}}\\
A_{\overline k k}A_{kk}^{-1}&A_{\overline k\overline k}-A_{\overline k k}A_{kk}^{-1}A_{k\overline{k}}
\end{bmatrix}
\begin{bmatrix}
u\\y
\end{bmatrix}=
\begin{bmatrix}
x\\v
\end{bmatrix},
\]
And conversely. Of course all this assumes that \(A_{kk}\) is non-singular. The following results are true for an arbitrary index set \(k\) and its complement
\(\overline k\).


1. $A=\mathbf{inv}_k(\mathbf{inv}_k(A))$
2. $\mathbf{inv}_k(A)=(\mathbf{inv}_{\overline k}(A))^{-1}$
3. $\mathbf{inv}_{\overline k}(\mathbf{inv}_k(A))=A^{-1}$
4. $\mathbf{inv}_{\ell}(\mathbf{inv}_k(A))=\mathbf{inv}_{k}(\mathbf{inv}_\ell(A))$
5. $\mathbf{inv}_{k}(A)=\mathbf{inv}_{\overline k}(A^{-1})$
6. $(\mathbf{inv}_{k}(A))^{-1}=\mathbf{inv}_{k}(A^{-1})$

#Timing

```{r timer}
set.seed(12345)
b <- crossprod(matrix(rnorm(1000000),1000,1000))/1000
```
```{r timing, echo = FALSE, cache = TRUE}
sweepTimer <- function (howMany = 5, whichPart = 1:500) {
	print ("element sweeping in pure R")
	gc()
	for (i in 1:howMany) {
		print (system.time(rSweep(b, 1:500)))
	}
	print ("element sweeping in R and fortran from R")
	gc()
	for (i in 1:howMany) {
		print (system.time(fSweep(b, 1:500)))
	}
	print ("element sweeping in R and C from R and fortran from C")
	gc()
	for (i in 1:howMany) {
		print (system.time(cSweep(b, 1:500)))
	}
	print ("just solve it")
	gc()
	for (i in 1:howMany) {
		print (system.time(solve(b)))
	}
}

sweepTimer()
```

#Code

##R Code

```{r display_code, eval = FALSE}

rSweep <- function (a, indi) {
for (j in indi) {
    pv <- a[j, j]
    if (pv == 0) next ()
    pr <- a[j, -j]
    pc <- a[-j, j]
    a[j, j] <- -1 / pv
    a[j, -j] <- pr / pv
    a[-j, j] <- pc / pv
    a[-j, -j] <- a[-j, -j] - outer(pc, pr) / pv
    }
return (a)
}

fSweep <- function (b , ind, eps = 1e-10, trianin = FALSE, trianout = FALSE) {
	if (trianin) {
		m <- length (b)
		n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
		a <- b
	} else {
		n <- nrow (b)
		m <- (n * n + n) / 2
		a <- b[upper.tri (b, diag = TRUE)]
	}
	for (k in ind) {
		a <- .Fortran("gsweep",
			diago = as.double (a[cumsum (1:n)]),
			trian = as.double (a),
			k = as.integer (k),
			l = as.integer (0),
			m = as.integer (m),
			n = as.integer (n),
			e = as.double (eps),
			ifault = as.integer (0))$trian
		}
	if (trianout) {
		return (a)
	} else {
		return (tri2sym (a))
	}
}

cSweep <- function (b , ind, eps = 1e-10, trianin = FALSE, trianout = FALSE) {
	if (trianin) {
		m <- length (b)
		n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
		a <- b
	} else {
		n <- nrow (b)
		m <- (n * n + n) / 2
		a <- b[upper.tri (b, diag = TRUE)]
	}
	a <- .C("hsweep",
		diago = as.double (a[cumsum (1:n)]),
		trian = as.double (a),
		n = as.integer (n),
		m = as.integer (m),
		ind = as.integer (ind),
		npiv = as.integer (length (ind)),
		eps = as.double (eps))$trian
	if (trianout) {
		return (a)
	} else {
		return (tri2sym (a))
	}
}

tri2sym <- function (a) {
	m <- length (a)
	n <- as.integer ( (sqrt (1 + 8 * m) - 1) / 2)
	b <- matrix (0, n, n)
	b[upper.tri (b, diag = TRUE)] <- a
	b <- b + t(b)
	diag (b) <- diag (b) / 2
	return (b)
}
```

##C code

```{r display_c_code, eval = FALSE, engine = "Rcpp"}
void gsweep_(double *, double *, int *, int *, int *, int *, double *, int *);

void
hsweep(double *s, double *trian, int *n, int *m, int *ind, int *np, double *eps) {
	int ifault = 0, l = 0;
	for (int i = 0; i < *np; i++) {
		gsweep_ (s, trian, ind + i, &l, m, n, eps, &ifault);
	}
```

#NEWS

#References
