---
title: "Computing and Fitting Monotone Splines"
author: "Jan de Leeuw"
date: "Version 10, March 17, 2017"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
fontsize: 12pt
graphics: yes
bibliography: splines.bib
abstract: An introduction to spline functions and B-splines, and specifically to monotone
  spline functions -- with code in R and C.
---
<style type="text/css">

body{ /* Normal  */
   font-size: 18px;
}
td {  /* Table  */
   font-size: 18px;
}
h1 { /* Header 1 */
 font-size: 28px;
 color: DarkBlue;
}
h2 { /* Header 2 */
 font-size: 22px;
 color: DarkBlue;
}
h3 { /* Header 3 */
 font-size: 18px;
 color: DarkBlue;
}
code.r{ /* Code block */
  font-size: 18px;
}
pre { /* Code block */
  font-size: 18px
}
</style>

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

```{r function_code, echo = FALSE}
source("lowSpline.R")
source("gslSpline.R")
source("splineBasis.R")
source("bispline.R")
```
```{r packages, echo = FALSE}
options (digits = 10) 
library (captioner)
library (lsei)
```
```{r equation_numbering, echo = FALSE}
figure_nums <- captioner (prefix = "Figure")
figure_nums (name = "rectangle_conf", caption = "Rectangle Data, Configuration, Original Left, Rounded Right", display = FALSE)
```
Note: This is a working paper which will be expanded/updated frequently. All suggestions for improvement are welcome. The directory [gifi.stat.ucla.edu/splines](http://gifi.stat.ucla.edu/splines) has a pdf version, the complete Rmd file with all code chunks, the bib file, 
and the R source code.

#Introduction

A *spline function* of *order* $k\geq 1$, or *degree* $k-1$, on a strictly increasing doubly infinite sequence $\{z_j\}$  of *knots*, 

* is a polynomial of degree at most $k-1$ between each pair of consecutive knots, and 
* the pieces are joined at the knots in such a way that the spline is everywhere $k-2$ times continuously differentiable. 

For convenience we assume $\lim_{j\rightarrow-\infty} z_j=-\infty$ and $\lim_{j\rightarrow+\infty} z_j=+\infty$.

The smoothness requirement needs some qualification. If $k=2$ we understand it as requiring continuity of the spline function, if $k=1$ it does not require anything. 

Thus a spline function of order one is a step function which steps form one level to the next at each knot. A spline of order two
is piecewise linear, which the pieces joined at the knots so that the spline function is continuous. Order three means piecewise quadratic and continuously differentiable at the knots. And so on. 

Note that we only consider spline functions with *simple knots* in this paper. The definition can be extended to the case in which consecutive knots are equal. i.e. knots have different *multiplicities*,  but this leads to complications which are not needed for our purposes. 

#Basic Splines

Alternatively, a spline function of order $k$ can be defined as a linear combination of *B-splines* of order $k$ on the same knot sequence. A B-spline, or basic spline, is a 
spline function that is positive in $k$ consecutive intervals and zero everywhere else. B-splines can be defined in various ways, using the piecewise polynomial definition,
the divided differences definition, or the recursive definition. The recursive definition is the most convenient one for computational purposes, and that is the one we use.

If we have $m$ knots we can define $m-k$ B-splines of order $k$, each covering $k$ consecutive intervals. At most $k$ different B-splines are non-zero at each point. B-spline
$j$ is zero for all $x\leq z_j$, then is positive for all $z_{j}<x<z_{j+k)}$, and zero again for $x>z_{j+k}$.

A key result in the theory of B-splines is that the different B-splines are linearly independent and form a basis for the linear space of spline functions (of a given order and knot sequence). There are two popular ways of normalizing B-splines. The $\mathcal{N}$-spline $\mathcal{N}_{kj}$, i.e. the normalized B-spline $j$ or order $k$, satisfies
$$
\sum_{\ell=j}^{j+n-k-1}\mathcal{N}_{k\ell}(x)=1
$$
for all $x$ and for all $j=1,\cdots,n-k$.

Alternatively we can normalize to $\mathcal{M}$-splines, for which
$$
\int_{-\infty}^{+\infty}\mathcal{M}_{kj}(x)dx=\int_{z_j}^{z_{j+k}}\mathcal{M}_{kj}(x)dx=1.
$$
We now illustrate normalized B-splines in detail, using nine knots.
```{r knot_sequence}
knots <- c(-2,-1,0,.1,.7,.9,1,2,3)
```
We evaluate our B-splines at $x_i$, equally spaced with steps $.001$ on the interval $[0.05,0.95]$.
```{r arguments}
x <- seq(0.05, 0.95, by = .001)
```
Note that all $x_i$ are in the unit interval, which means that for this problem the three smallest and the three largest knots are *external knots*, while the three middle knots
are *internal knots*. If we want our $n$ splines to have $k$ nonzero values at our $n$ arguments $x_i$, we need $k$ external knots smaller than `min (x)` and $k$ external knots larger than `max(x)`. . In @wang_yan_17  and  @sinha the *boundary knots* are chosen as $k$ copies of the minimum and $k$ copies of the maximum of the $x_i$, in @deleeuw_E_15d we use $k$ copies of `min (c (knots, x))` and `max( c (knots, x))`. But both these choices violate the simple knots restriction used in this paper.

The file `lowSpline.R` has the three functions `ZbSpline()`, `LbSpline()`, and `QbSpline` that use the explicit piecewise polynomial representation to compute B-splines or
orders one, two, and three.

In our example there are eight B-splines or order one. Only four are not identically zero. The normalized B-splines $\mathcal{N}_{kj}$ are equal to one in the half-open interval $[z_j,z_{j+1})$ and zero outside the interval. Note that B-splines of order one are right-continuous. The corresponding $\mathcal{M}_{kj}$ have a value equal to the
reciprocal of the length of the interval.

```{r zbspline}
par (mfrow = c (1,2))
for (k in 1:8) {
  y <- ZbSpline (x, knots, k = k)
  if (all (y == 0)) next
  stepPlotter (x, y, knots, xlab = "x")
  for (i in 1:9) abline (v = knots[i])
}
```


B-splines of order two are *hat functions*, with an increasing linear function from $z_j$ to $z_{j+1}$ and a decreasing linear function from $z_{j+1}$ to $z_{j+2}$. There are
seven B-splines of order $k=2$ in our example, but only the middle five are non-zero on the range $[0.05,0.95]$.
```{r lbspline}
par (mfrow = c (1,2))
for (k in 1:7) {
  y <- LbSpline (x, knots, k = k)
  if (all (y == 0)) next
  plot (x, y, type = "l", col = "RED", lwd = 3)
  for (i in 1:9) abline (v = knots[i])
}

```

B-splines of order three are *hat functions*, with an increasing linear part from $z_j$ to $z_{j+1}$ and a decreasing linear part from $z_{j+1}$ to $z_{j+2}$. There are
six B-splines of order three in our example, and all six are non-zero on the range $[0.05,0.95]$.

```{r qbspline}
par (mfrow = c (1,2))
for (k in 1:6) {
  y <- QbSpline (x, knots, k = k)
  if (all (y == 0)) next
  plot (x, y, type = "l", col = "RED", lwd = 3)
  for (i in 1:9) abline (v = knots[i])
}
```
##Derivatives of B-splines

##Integrals of B-splines


#Example

```{r births_b_splines_plot}
births <- scan("http://robjhyndman.com/tsdldata/data/nybirths.dat")
birthsts <- ts(births, frequency=12, start=c(1946,1))
plot(birthsts, col = "RED", lwd = 2)
```
```{r births_fit}
knots <- 12 * (-3:17) + .5
h <- bisplineBasis(1:168,2,knots)
b <- h$bbasis
nb <- ncol(b)
par(mfrow=c(1,2))
for (i in 1:nb) {
  plot(1:168, b[, i], type = "l", col = "RED", lwd = 3, xlab = "month", ylab = "b-spline")
  for (y in knots) abline (v=y)
}
```
```{r}
u<-lm.fit(b, births)
plot(1:168, births, type = "l", col = "RED", lwd = 3, xlab = "months", ylab = "births")
lines (1:168, b%*%u$coefficients, col = "BLUE", lwd = 3)
for (y in knots) abline (v=y)
```

`r sum ((births - b%*%u$coefficients)^2)`

#B-Splines with monotone weights

$$
\min_{\beta_1\leq\beta_2\leq\cdots\leq\beta_p}\mathbf{SSQ}(y-X\beta)
$$

```{r}
d<-matrix(0,nb-1,nb)
diag(d)=-1
d[outer(1:(nb-1),1:nb,function(i,j) (j - i) == 1)]<-1
u<-lsi(b,births,e=d,f=rep(0,nb-1))
plot(1:168, births, type = "l", col = "RED", lwd = 3, xlab = "months", ylab = "births")
lines (1:168, b%*%u, col = "BLUE", lwd = 3)
for (y in knots) abline (v=y)
```

`r sum ((births - b%*%u)^2)`

```{r}
h3 <- bisplineBasis(1:168,3,knots[1:13],lowknot = .5, highknot = 168.5)
b3 <- h3$bbasis
nb <- ncol(b3)
d<-matrix(0,nb-1,nb)
diag(d)=-1
d[outer(1:(nb-1),1:nb,function(i,j) (j - i) == 1)]<-1
u<-lsi(b3,births,e=d,f=rep(0,nb-1))
plot(1:168, births, type = "l", col = "RED", lwd = 2, xlab = "months", ylab = "births")
lines (1:168, b3%*%u)
```

#B-Splines with monotone values


```{r}
mregnnM <- function (x, y) {
  k <- qr.Q(qr(x))
  u <- drop (crossprod(k, y))
  v <- -t(diff(k))
  lb <- nnls(v, u)$x
  xb <- drop(k %*% (u - v%*% lb))
  return (list(xb = xb,lb = lb,f = sum((y - xb) ^ 2)))
}
```
```{r}
u<-mregnnM(b, births)
plot(1:168, births, type = "l", col = "RED", lwd = 2, xlab = "months", ylab = "births")
lines (1:168, u$xb)
```


`r sum ((births - u$xb)^2)`

$$
\min_{x_1'\beta\leq x_2'\beta\leq\cdots\leq x_n'\beta}\mathbf{SSQ}(y-X\beta)
$$



#I-splines

$$
\min_{\beta\geq 0}\mathbf{SSQ}(y-Z\beta)
$$

```{r births_isplines_plot}
par(mfrow=c(1,2))
h <- bisplineBasis(1:168,2,knots, norm = 1)
g <- h$ibasis
ng <- ncol(g)
for (i in 1:ng) {
  plot(1:168, g[, i], type = "l", col = "RED", lwd = 2, xlab = "month", ylab = "i-spline")
}
```

```{r births_i_splines_fit}
u<-nnls (g,births)
plot(1:168, births, type = "l", col = "RED", lwd = 2, xlab = "months", ylab = "births")
lines (1:168, g%*%u$x)
```

`r sum ((births - g%*%u$x)^2)`

#Regression Example

```{r example}
data(neumann, package = "homals")
neumann
```

#Appendix: Code

##R code

###lowspline.R

```{r file_auxilary, code = readLines("lowspline.R")}
```

###gslSpline.R

```{r file_auxilary, code = readLines("gslSpline.R")}
```

###splinebasis.R

```{r file_auxilary, code = readLines("splinebasis.R")}
```

###bispline.R

```{r file_auxilary, code = readLines("bispline.R")}
```

##C code

###gslSpline.c

```{c file_auxilary, code = readLines("gslSpline.c")}
```

###splinebasis.c

```{c file_auxilary, code = readLines("splinebasis.c")}
```

###bispline.c

```{c file_auxilary, code = readLines("bispline.c")}
```

